\bsection[META-NET Mitglieder -- META-NET Members]{META-NET Mitglieder --- META-NET Members}
\label{metanetmembers}

\small
\begin{longtable}{llp{105mm}}
  Belgien & \textcolor{grey1}{Belgium} & Computational Linguistics and Psycholinguistics Research Centre, Univ. of Antwerp: Walter Daelemans\\ \addlinespace
  & & Centre for Proc. Speech and Images, Univ. of Leuven: Dirk van Compernolle \\ \addlinespace
  Bulgarien & \textcolor{grey1}{Bulgaria} & Inst. for Bulgarian Lang., Bulgarian Academy of Sciences: Svetla Koeva \\ \addlinespace
  Deutschland & \textcolor{grey1}{Germany} & DFKI (German Research Centre for Artificial Intelligence): Hans Uszkoreit, Georg Rehm\\ \addlinespace
  & & Human Lang. Technology and Pattern Recognition, RWTH Aachen Univ.: Hermann Ney \\ \addlinespace
  & & Dept. of Computational Linguistics, Saarland Univ.: Manfred Pinkal\\ \addlinespace Dänemark &  \textcolor{grey1}{Denmark} & Centre for Lang. Technology, Univ. of Copenhagen: Bolette Sandford Pedersen, Bente Maegaard\\ \addlinespace
  Estland & \textcolor{grey1}{Estonia} & Inst. of Computer Science, Univ. of Tartu: Tiit Roosmaa, Kadri Vider\\ \addlinespace
  Finnland & \textcolor{grey1}{Finland} & Computational Cognitive Systems Research Group, Aalto Univ.: Timo Honkela\\ \addlinespace
  & & Dept. of General Linguistics, Univ. of Helsinki: Kimmo Koskenniemi, Krister Linden \\ \addlinespace
  Frankreich & \textcolor{grey1}{France} & Centre National de la Recherche Scientifique, Laboratoire d'Informatique pour la Mécanique et les Sciences de l'Ingénieur: Joseph Mariani \\ \addlinespace
  & & Evaluations and Lang. Resources Distribution Agency: Khalid Choukri\\ \addlinespace 
  Griechenland & \textcolor{grey1}{Greece} & R.C. “Athena”, Inst. for Language and Speech Processing: Stelios Piperidis\\ \addlinespace
  Großbritannien & \textcolor{grey1}{UK} & 
  School of Computer Science, Univ. of Manchester: Sophia Ananiadou \\ \addlinespace 
  & & Inst. for Lang., Cognition and Computation, Center for Speech Technology Research, Univ. of Edinburgh: Steve Renals \\ \addlinespace 
  & & Research Inst. of Informatics and Lang. Proc., Univ. of Wolverhampton: Ruslan Mitkov \\ \addlinespace 
  Irland & \textcolor{grey1}{Ireland} & School of Computing, Dublin City Univ.: Josef van Genabith\\ \addlinespace
  Island & \textcolor{grey1}{Iceland} & School of Humanities, Univ. of Iceland: Eirikur Rögnvaldsson\\ \addlinespace
  Italien & \textcolor{grey1}{Italy} & Consiglio Nazionale delle Ricerche, Istituto di Linguistica Computazionale “Antonio Zampolli”: Nicoletta Calzolari\\ \addlinespace
  & & Human Language Technology Research Unit, Fondazione Bruno Kessler: Bernardo Magnini\\ \addlinespace 
  Kroatien & \textcolor{grey1}{Croatia} & Inst. of Linguistics, Faculty of Humanities and Social Science, Univ. of Zagreb: Marko Tadić \\ \addlinespace
  Lettland & \textcolor{grey1}{Latvia} & Tilde: Andrejs Vasiljevs\\ \addlinespace 
  & & Inst. of Mathematics and Computer Science, Univ. of Latvia: Inguna Skadina\\ \addlinespace
  Litauen & \textcolor{grey1}{Lithuania} & Inst. of the Lithuanian Lang.: Jolanta Zabarskaitė\\ \addlinespace
  Luxemburg & \textcolor{grey1}{Luxembourg} & Arax Ltd.: Vartkes Goetcherian\\ \addlinespace
  Malta & \textcolor{grey1}{Malta} & Dept. Intelligent Computer Systems, Univ. of Malta: Mike Rosner\\ \addlinespace
  Niederlande & \textcolor{grey1}{Netherlands} & Utrecht Inst. of Linguistics, Utrecht Univ.: Jan Odijk\\ \addlinespace 
  & & Computational Linguistics, Univ. of Groningen: Gertjan van Noord\\ \addlinespace
  Norwegen & \textcolor{grey1}{Norway} & Dept. of Linguistic, Univ. of Bergen: Koenraad De Smedt\\ \addlinespace 
  & & Dept. of Informatics, Lang. Technology Group, Univ. of Oslo: Stephan Oepen \\ \addlinespace
  Österreich & \textcolor{grey1}{Austria} & Zentrum für Translationswissenschaft, Universität Wien: Gerhard Budin\\ \addlinespace 
  Polen & \textcolor{grey1}{Poland} & Inst. of Computer Science, Polish Academy of Sciences: Adam Przepiórkowski, Maciej Ogrodniczuk \\ \addlinespace
  & & Univ. of Łódź: Barbara Lewandowska-Tomaszczyk, Piotr Pęzik\\ \addlinespace
  & & Dept. of Computer Linguistics and Artificial Intelligence, Adam Mickiewicz Univ.: Zygmunt Vetulani \\ \addlinespace
  Portugal & \textcolor{grey1}{Portugal} & Dept. of Informatics, Univ. of Lisbon: Antonio Branco\\ \addlinespace
  & & Spoken Lang. Systems Lab., Inst. for Systems Engineering and Computers: Isabel Trancoso \\ \addlinespace
  Rumänien & \textcolor{grey1}{Romania} & Research Inst. for Artiﬁcial Intelligence, Romanian Academy of Sciences: Dan Tuﬁș \\ \addlinespace
  & & Faculty of Computer Science, Univ. Alexandru Ioan Cuza of Iași: Dan Cristea \\ \addlinespace
  Schweden & \textcolor{grey1}{Sweden} & Dept. of Swedish Lang., Univ. of Gothenburg: Lars Borin \\ \addlinespace 
  Schweiz & \textcolor{grey1}{Switzerland} & Idiap Research Inst.: Hervé Bourlard \\ \addlinespace 
  Serbien & \textcolor{grey1}{Serbia} & University of Belgrade, Faculty of Mathematics, Duško Vitas, Cvetana Krstev, Ivan Obradović \\ \addlinespace
  & & Pupin Inst.: Sanja Vranes \\ \addlinespace  
  Slowakei & \textcolor{grey1}{Slovakia} & Ľudovít Štúr Institute of Linguistics, Slovak Academy of Sciences: Radovan Garabík \\ \addlinespace 
  Slowenien & \textcolor{grey1}{Slovenia} & Jozef Stefan Inst.: Marko Grobelnik \\ \addlinespace 
  Spanien & \textcolor{grey1}{Spain} & Barcelona Media: Toni Badia, Maite Melero \\ \addlinespace 
  & & Institut Universitari de Lingüistica Aplicada, Univ. Pompeu Fabra: Núria Bel \\ \addlinespace 
  & & Aholab Signal Proc. Lab., Univ. of the Basque Country: Inma Hernaez Rioja \\ \addlinespace 
  & & Center for Language and Speech Technologies and Applications, Universitat Politècnica de Catalunya:  Asunción Moreno \\ \addlinespace 
  & & Dept. of Signal Proc. and Communications, Univ. of Vigo: Carmen García Mateo \\ \addlinespace 
  Tschechien & \textcolor{grey1}{Czech Republic} & Inst. of Formal and Applied Linguistics, Charles Univ. in Prague: Jan Hajic \\ \addlinespace
  Ungarn & \textcolor{grey1}{Hungary} & Research Inst. for Linguistics, Hungarian Academy of Sciences: Tamás Váradi\\  \addlinespace
  & & Dept. of Telecommunications and Media Informatics, Budapest Univ. of Technology and Economics: Géza Németh and Gábor Olaszy\\ \addlinespace
  Zypern & \textcolor{grey1}{Cyprus} & Lang. Centre, School of Humanities: Jack Burston
\end{longtable}
\normalsize
