%                                     MMMMMMMMM                                         
%                                                                             
%  MMO    MM   MMMMMM  MMMMMMM   MM    MMMMMMMM   MMD   MM  MMMMMMM MMMMMMM   
%  MMM   MMM   MM        MM     ?MMM              MMM$  MM  MM         MM     
%  MMMM 7MMM   MM        MM     MM8M    MMMMMMM   MMMMD MM  MM         MM     
%  MM MMMMMM   MMMMMM    MM    MM  MM             MM MMDMM  MMMMMM     MM     
%  MM  MM MM   MM        MM    MMMMMM             MM  MMMM  MM         MM     
%  MM     MM   MMMMMM    MM   MM    MM            MM   MMM  MMMMMMM    MM
%
%
%          - META-NET Language Whitepaper | Catalan Metadata -
% 
% ----------------------------------------------------------------------------

\usepackage{polyglossia}
\setotherlanguages{catalan,english}


\title{La llengua catalana a l'era digital --- The Catalan Language\ \
\ \ in the\ \ \ \ Digital Age}


\spineTitle{The Catalan Language in the Digital Age --- La llengua catalana a l'era digital}

\subtitle{White Paper Series --- Sèrie de Llibres Blancs}

\author{
 Asunción Moreno\\
 Núria Bel\\ 
 Eva Revilla\\
 Emília Garcia\\
 Sisco Vallverdú
}

\authoraffiliation{
 Asunción Moreno~ {\small [UPC]}\\
 Núria Bel~{\small [UPF]}\\ 
 Eva Revilla~{\small [UPF]}\\
 Emília Garcia~{\small [UPC]}\\
 Sisco Vallverdú~{\small [UPC]}\\
 \footnotesize{[UPC]} ~ \small{Universitat Politècnica de Catalunya}\\
 \footnotesize{[UPF]} ~ \small{Universitat Pompeu Fabra}
}


\editors{
  Georg Rehm, Hans Uszkoreit\\(editors-catalan \textcolor{grey1}{editors})
}

% Text in left column on backside of the cover
\SpineLText{\selectlanguage{english}%
  In everyday communication, Europe’s citizens, business partners and politicians are inevitably confronted with language barriers.  
  Language technology has the potential to overcome these barriers and to provide innovative interfaces to technologies and knowledge. 
  This white paper presents the state of language technology support for the Catalan language. 
  It is part of a series that analyses the available language resources and technologies for 30~European languages. 
  The analysis was carried out by META-NET, a Network of Excellence funded by the European Commission.
  META-NET consists of 54 research centres in 33 countries, who cooperate with stakeholders from economy, government agencies, research organisations, non-governmental organisations, language communities and European universities. 
  META-NET’s vision is high-quality language technology for all European languages.
}

% Text in right column on backside of the cover
\SpineRText{\selectlanguage{catalan}%
En la comunicació quotidiana, els ciutadans europeus, socis de negocis i polítics s'enfronten inevitablement amb les barreres de l'idioma.
La tecnologia del llenguatge té el potencial per superar aquestes barreres i proporcionar interfícies innovadores a les tecnologies i coneixements.
Aquest llibre blanc presenta l'estat del suport de la tecnologica del llenguatge per al català.
És part d'una sèrie de llibres que analitza els recursos i tecnologies disponibles per 30~llengües europees.
L'anàlisi va ser realitzat per META-NET, una xarxa d'excel·lència finançada per la Comissió Europea.
META-NET es compon de 54 centres de recerca a 33 països, que col·laboren amb les parts interessades de l'economia, les agències governamentals, organitzacions de recerca, organitzacions no governamentals, comunitats lingüístiques i universitats europees.
La visió de META-NET és tecnologia del llenguatge d'alta qualitat per totes les llengües europees.
}


% Funding notice left column
\FundingLNotice{\selectlanguage{catalan} Els autors d'aquest document 
agraeixen als autors del Llibre Blanc en Alemany \cite{lwpgerman} el permís per 
reutilitzar fragments del seu document independents de la llengua.
  
  \bigskip
  El desenvolupament d'aquest document ha estat finançat pel Setè 
  Programa Marc i el Programa de Suport a Polítiques de les TIC 
  de la Comissió Europea amb els contractes T4ME (acord de 
  subvenció 249119), CESAR (acord de subvenció 271022), 
  METANET4U (acord de subvenció del 270 893) i META-NORD 
  (acord de subvenció 270899).}

% Funding notice right column
\FundingRNotice{\selectlanguage{english} The authors of this document
  are grateful to the authors of the White Paper on German \cite{lwpgerman} for
  permission to re-use selected language-independent materials from
  their document.
  
  \bigskip
  The development of this white paper has been funded by the Seventh
  Framework Programme and the ICT Policy Support Programme of the
  European Commission under the contracts T4ME (Grant Agreement
  249119), CESAR (Grant Agreement 271022), METANET4U (Grant Agreement
  270893) and META-NORD (Grant Agreement 270899).}
