
\small
\begin{longtable}{llp{105mm}}
  Belgicko & \textcolor{grey1}{Belgium} & Computational Linguistics and Psycholinguistics Research Centre, Univ. of Antwerp: Walter Daelemans\\ \addlinespace & & Centre for Proc. Speech and Images, Univ. of Leuven: Dirk van Compernolle \\ \addlinespace
  Bulharsko & \textcolor{grey1}{Bulgaria} & Inst. for Bulgarian Lang., Bulgarian Academy of Sciences: Svetla Koeva \\ \addlinespace
  Cyprus & \textcolor{grey1}{Cyprus} & Lang. Centre, School of Humanities: Jack Burston \\ \addlinespace
  Česká republika & \textcolor{grey1}{Czech Republic} & Inst. of Formal and Applied Linguistics, Charles Univ. in Prague: Jan Hajič \\ \addlinespace
  Dánsko &  \textcolor{grey1}{Denmark} & Centre for Lang. Technology, Univ. of Copenhagen: Bolette Sandford Pedersen, Bente Maegaard\\ \addlinespace
  Estónsko & \textcolor{grey1}{Estonia} & Inst. of Computer Science, Univ. of Tartu: Tiit Roosmaa\\ \addlinespace
  Fínsko & \textcolor{grey1}{Finland} & Computational Cognitive Systems Research Group, Aalto Univ.: Timo Honkela\\ \addlinespace  & & Dept. of General Linguistics, Univ. of Helsinki: Kimmo Koskenniemi, Krister Linden \\ \addlinespace
  Francúzsko & \textcolor{grey1}{France} & Centre National de la Recherche Scientifique, Laboratoire d'Informatique pour la Mécanique et les Sciences de l'Ingénieur: Joseph Mariani \\ \addlinespace  & & Evaluations and Lang. Resources Distribution Agency: Khalid Choukri\\ \addlinespace 
  Grécko & \textcolor{grey1}{Greece} & Inst. for Lang. and Speech Proc., R.C. “Athena”: Stelios Piperidis\\ \addlinespace
  Holandsko & \textcolor{grey1}{Netherlands} & Utrecht Inst. of Linguistics, Utrecht Univ.: Jan Odijk\\ \addlinespace   & & Computational Linguistics, Univ. of Groningen: Gertjan van Noord\\ \addlinespace
  Chorvátsko & \textcolor{grey1}{Croatia} & Inst. of Linguistics, Faculty of Humanities and Social Science, Univ. of Zagreb: Marko Tadić \\ \addlinespace
  Island & \textcolor{grey1}{Iceland} & School of Humanities, Univ. of Island: Eirikur Rögnvaldsson\\ \addlinespace
  Írsko & \textcolor{grey1}{Ireland} & School of Computing, Dublin City Univ.: Josef van Genabith\\ \addlinespace
  Litva & \textcolor{grey1}{Lithuania} & Inst. of the Lithuanian Lang.: Jolanta Zabarskaitė\\ \addlinespace
  Lotyšsko & \textcolor{grey1}{Latvia} & Tilde: Andrejs Vasiļjevs\\ \addlinespace   & & Inst. of Mathematics and Computer Science, Univ. of Latvia: Inguna Skadina\\ \addlinespace
  Luxembursko & \textcolor{grey1}{Luxembourg} & Arax Ltd.: Vartkes Goetcherian\\ \addlinespace
  Maďarsko & \textcolor{grey1}{Hungary} & Research Inst. for Linguistics, Hungarian Academy of Sciences: Tamás Váradi\\  \addlinespace  & & Dept. of Telecommunications and Media Informatics, Budapest Univ. of Technology and Economics: Géza Németh and Gábor Olaszy\\ \addlinespace
  Malta & \textcolor{grey1}{Malta} & Dept. Intelligent Computer Systems, Univ. of Malta: Mike Rosner\\ \addlinespace
  Nemecko & \textcolor{grey1}{Germany} & DFKI (German Research Centre for Artificial Intelligence): Hans Uszkoreit, Georg Rehm\\ \addlinespace  & & Human Lang. Technology and Pattern Recognition, RWTH Aachen Univ.: Hermann Ney \\ \addlinespace  & & Dept. of Computational Linguistics, Saarland Univ.: Manfred Pinkal\\ \addlinespace 
  Nórsko & \textcolor{grey1}{Norway} & Dept. of Linguistic, Univ. of Bergen: Koenraad De Smedt\\ \addlinespace   & & Dept. of Informatics, Lang. Technology Group, Univ. of Oslo: Stephan Oepen \\ \addlinespace
  Poľsko & \textcolor{grey1}{Poland} & Inst. of Computer Science, Polish Academy of Sciences: Adam Przepiórkowski, Maciej Ogrodniczuk \\ \addlinespace  & & Univ. of Łódź: Barbara Lewandowska-Tomaszczyk, Piotr Pęzik\\ \addlinespace  & & Dept. of Computer Linguistics and Artificial Intelligence, Adam Mickiewicz Univ.: Zygmunt Vetulani \\ \addlinespace
  Portugalsko & \textcolor{grey1}{Portugal} & Dept. of Informatics, Univ. of Lisbon: Antonio Branco\\ \addlinespace  & & Spoken Lang. Systems Lab., Inst. for Systems Engineering and Computers: Isabel Trancoso \\ \addlinespace
  Rakúsko & \textcolor{grey1}{Austria} & Zentrum für Translationswissenschaft, Universität Wien: Gerhard Budin\\ \addlinespace 
  Rumunsko & \textcolor{grey1}{Romania} & Research Inst. for Artificial Intelligence, Romanian Academy of Sciences: Dan Tufiș \\ \addlinespace  & & Faculty of Computer Science, Univ. Alexandru Ioan Cuza: Dan Cristea \\ \addlinespace
  Slovensko & \textcolor{grey1}{Slovakia} & Ľ. Štúr Institute of Linguistics, Slovak Academy of Sciences: Radovan Garabík \\ \addlinespace 
  Slovinsko & \textcolor{grey1}{Slovenia} & Jožef Stefan Inst.: Marko Grobelnik \\ \addlinespace 
      \begin{minipage}[t]{0.2\columnwidth}%
      Spojené kráľovstvo \\
      Veľkej Británie a \\
      Severného Írska 
      \end{minipage}
            & \begin{minipage}[t]{0.15\columnwidth}%
                \hfill \\
                \textcolor{grey1}{UK} 
              \end{minipage}
            & Inst. for Lang., Cognition and Computation, Center for Speech Technology Research, Univ. of Edinburgh: Steve Renals \\ \addlinespace   & & Research Inst. of Informatics and Lang. Proc., Univ. of Wolverhampton: Ruslan Mitkov \\ \addlinespace   & & School of Computer Science, Univ. of Manchester: Sophia Ananiandou  \\ \addlinespace 
  Srbsko & \textcolor{grey1}{Serbia} & Faculty of Mathematics, Belgrade Univ.: Duško Vitas, Cvetana Krstev, Ivan Obradović \\ \addlinespace  & & Pupin Inst.: Sanja Vraneš \\ \addlinespace  
  Španielsko & \textcolor{grey1}{Spain} & Barcelona Media: Toni Badia \\ \addlinespace   & & Institut Universitari de Lingüistica Aplicada, Univ. Pompeu Fabra: Núria Bel \\ \addlinespace   & & Aholab Signal Proc. Lab., Univ. of the Basque Country: Inma Hernaez Rioja \\ \addlinespace   & & Center for Lang. and Speech Technologies and Applications, Technical Univ. of Catalonia: Asunción Moreno \\ \addlinespace   & & Dept. of Signal Proc. and Communications, Univ. of Vigo: Carmen García Mateo \\ \addlinespace 
  Švajčiarsko & \textcolor{grey1}{Switzerland} & Idiap Research Inst.: Hervé Bourlard \\ \addlinespace 
  Švédsko & \textcolor{grey1}{Sweden} & Dept. of Swedish Lang., Univ. of Gothenburg: Lars Borin \\ \addlinespace 
  Taliansko & \textcolor{grey1}{Italy} & Consiglio Nazionale Ricerche, Istituto di Linguistica Computazionale “Antonio Zampolli”: Nicoletta Calzolari\\ \addlinespace  & & Human Lang. Technology, Fondazione Bruno Kessler: Bernardo Magnini
\end{longtable}
\normalsize

